---
title: "LIPIDIFy: A Complete Lipidomics Analysis Workflow"
author:
  - name: "Package Authors"
package: LIPIDIFy
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{LIPIDIFy: A Complete Lipidomics Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

**LIPIDIFy** is an R package providing a complete, interactive analysis
pipeline for lipidomics mass-spectrometry data. It includes:

- Automatic lipid classification by class, type, and fatty-acid saturation
- Multiple normalization strategies with visual comparison
- Differential abundance analysis (limma / EdgeR)
- Lipid-class enrichment analysis (fgsea)
- An interactive Shiny dashboard for point-and-click analysis
- Automated HTML/PDF report generation

This vignette demonstrates the full command-line workflow using a built-in
example dataset.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("LIPIDIFy")
```

# Quick Start — Interactive App

The fastest way to explore LIPIDIFy is via the Shiny dashboard:

```{r launch-app, eval=FALSE}
library(LIPIDIFy)
launch_lipidomics_app()
```

The rest of this vignette demonstrates the same analysis programmatically.

# Load the Example Dataset

```{r load-data}
library(LIPIDIFy)

raw_df   <- generate_example_data()
data_obj <- load_lipidomics_data_from_df(raw_df)

cat("Samples :", nrow(data_obj$numeric_data), "\n")
cat("Lipids  :", ncol(data_obj$numeric_data), "\n")
cat("Groups  :", paste(unique(data_obj$metadata$`Sample Group`),
                       collapse = ", "), "\n")
```

# Lipid Classification

```{r classify}
lipid_names    <- colnames(data_obj$numeric_data)
classification <- classify_lipids(lipid_names)

head(classification)
table(classification$LipidGroup)
table(classification$Saturation)
```

# Normalization

## Available methods

```{r norm-methods}
get_normalization_methods()
```

Descriptions of each method are available via `get_normalization_descriptions()`.

> **Note on Quantile Normalization:** after quantile normalization all
> per-sample boxplots look nearly identical. This is the *expected, correct*
> behaviour — the method forces every sample to share the same intensity
> distribution.

> **Note on Median vs Mean:** on symmetric (log-normal) lipidomics data the
> two methods produce visually similar results. On skewed data they differ.
> This is expected and is not a bug.

## Apply a normalization pipeline

```{r normalize}
norm_data <- apply_normalizations(
  data_obj$numeric_data,
  methods = c("TIC", "Log2")
)

# Quick sanity check: row sums after TIC should be equal
tic_only <- normalize_tic(data_obj$numeric_data)
round(rowSums(tic_only), 2)  # all values should be identical
```

## Visualize raw vs normalized

```{r viz-raw, fig.cap="Raw data distribution by sample"}
raw_list  <- list(numeric_data = data_obj$numeric_data)
norm_list <- list(numeric_data = norm_data,
                  metadata     = data_obj$metadata)

p_raw  <- visualize_raw_data_improved(raw_list,  "boxplot", "sample")
p_norm <- visualize_raw_data_improved(norm_list, "boxplot", "sample",
                                      metadata     = data_obj$metadata,
                                      group_column = "Sample Group")
print(p_raw)
print(p_norm)
```

# PCA

```{r pca, fig.cap="PCA coloured by sample group with 95% confidence ellipses"}
pca_res <- perform_pca(norm_data, data_obj$metadata, "Sample Group")

create_pca_plot_with_ellipses(
  pca_data           = pca_res$pca_data,
  variance_explained = pca_res$variance_explained,
  ellipse_type       = "confidence",
  show_sample_labels = TRUE
)
```

# Differential Analysis

```{r diff-analysis}
diff_res <- perform_differential_analysis(
  data_matrix    = norm_data,
  metadata       = data_obj$metadata,
  group_column   = "Sample Group",
  contrasts_list = c("GroupB - GroupA", "GroupC - GroupA"),
  method         = "limma"
)

# Summary
for (nm in names(diff_res$results)) {
  res   <- diff_res$results[[nm]]
  n_sig <- sum(res$adj.P.Val < 0.05, na.rm = TRUE)
  message(nm, ": ", n_sig, " significant lipids")
}
```

## Volcano plot

```{r volcano, fig.cap="Volcano plot for GroupB vs GroupA"}
create_volcano_plot_labeled(
  results         = diff_res$results[["GroupB - GroupA"]],
  title           = "Volcano: GroupB vs GroupA",
  logfc_threshold = 1,
  pval_threshold  = 0.05,
  top_labels      = 10,
  classification_data = classification,
  color_by        = "LipidGroup"
)
```

# Lipid Expression

```{r expression, fig.cap="Per-sample abundance of selected lipids, sorted by group"}
# Select the top 3 most significantly changed lipids
top_lipids <- utils::head(
  rownames(diff_res$results[["GroupB - GroupA"]])[
    order(diff_res$results[["GroupB - GroupA"]]$adj.P.Val)
  ], 3
)

plots <- create_lipid_expression_barplot(
  data_matrix    = norm_data,
  metadata       = data_obj$metadata,
  selected_lipids = top_lipids,
  group_column   = "Sample Group"
)

# Print first plot (samples sorted by group automatically)
print(plots[[1]])
```

# Enrichment Analysis

```{r enrichment}
enrich_res <- perform_enrichment_analysis(
  results_list      = diff_res$results,
  classification_data = classification,
  min_set_size      = 3,
  max_set_size      = 500
)

# Show available results
str(enrich_res, max.level = 2)
```

# Session Information

```{r session-info}
sessionInfo()
```
