---
title: "LIPIDIFy: A Complete Lipidomics Analysis Workflow"
author: "Fayrouz Hammal"
date: "`r Sys.Date()`"
package: LIPIDIFy
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{LIPIDIFy: A Complete Lipidomics Analysis Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 9,
  fig.height = 5,
  fig.align = "center",
  warning   = FALSE,
  message   = FALSE
)
```

## Overview

**LIPIDIFy** provides an end-to-end pipeline for mass-spectrometry lipidomics data:

| Step | What it does | Key function(s) |
|------|-------------|-----------------|
| 1 | Load data | `generate_example_data()`, `load_lipidomics_data_from_df()` |
| 2 | Classify lipids | `classify_lipids()` |
| 3 | Visualize raw data | `visualize_raw_data_improved()` |
| 4 | Compare and apply normalization | `apply_normalizations()`, `create_pipeline_plot()` |
| 5 | Quality control (PCA) | `perform_pca()`, `create_pca_plot_with_ellipses()` |
| 6 | Differential analysis | `perform_differential_analysis()` |
| 7 | Visualize results | `create_volcano_plot_labeled()`, `create_heatmap_robust()` |
| 8 | Lipid expression plots | `create_lipid_expression_barplot()` |
| 9 | Enrichment analysis | `perform_enrichment_analysis()` |

The same workflow is available without any coding via the Shiny application:

```{r launch-app, eval=FALSE}
library(LIPIDIFy)
launch_lipidomics_app()
```

***

## Step 1 - Load the Example Dataset

`generate_example_data()` creates a reproducible synthetic dataset with
realistic lipid names (PC, PE, TG, SM, Cer, etc.) and four sample groups
(GroupA, GroupB, GroupC, GroupD; 5 replicates each; 301 lipids total).

```{r load-data}
library(LIPIDIFy)

# Generate the built-in synthetic dataset (seed = 42 for reproducibility)
raw_df   <- generate_example_data()
data_obj <- load_lipidomics_data_from_df(raw_df)

cat("Samples  :", nrow(data_obj$numeric_data), "\n")
cat("Lipids   :", ncol(data_obj$numeric_data), "\n")
cat("Groups   :", paste(unique(data_obj$metadata$`Sample Group`),
                        collapse = ", "), "\n")
```

`load_lipidomics_data_from_df()` returns a **named list** with three elements:

- `$numeric_data` - matrix of lipid abundances (samples x lipids)
- `$metadata` - data frame of sample metadata
- `$data` - the original full data frame

```{r peek-data}
# Metadata of the first 4 samples
head(data_obj$metadata[, c("Sample Name", "Sample Group", "Tumour ID")], 4)

# First 3 lipids of the first 4 samples
data_obj$numeric_data[1:4, 1:3]
```

***

## Step 2 - Lipid Classification

`classify_lipids()` parses lipid names and assigns three classification levels:

- **LipidGroup** - broad category (e.g., Glycerophospholipids, Sphingolipids)
- **LipidType** - specific subclass (e.g., Phosphatidylcholine)
- **Saturation** - SFA (0 double bonds), MUFA (1), PUFA (> 1)

```{r classify}
lipid_names    <- colnames(data_obj$numeric_data)
classification <- classify_lipids(lipid_names)

head(classification, 8)
```

```{r classify-summary}
cat("Lipid groups:\n")
print(sort(table(classification$LipidGroup), decreasing = TRUE))

cat("\nFatty-acid saturation:\n")
print(table(classification$Saturation))
```

***

## Step 3 - Raw Data Visualization

Examining distributions **before normalization** helps detect outlier samples,
batch effects, and scaling differences.

```{r raw-boxplot, fig.cap="Per-sample intensity distributions before normalization. Boxes are coloured by group."}
raw_list <- list(numeric_data = data_obj$numeric_data,
                 metadata     = data_obj$metadata)

visualize_raw_data_improved(
  raw_list,
  plot_type    = "boxplot",
  view_mode    = "sample",
  metadata     = data_obj$metadata,
  group_column = "Sample Group"
)
```

```{r raw-density, fig.cap="Intensity density curves before normalization, one curve per sample."}
visualize_raw_data_improved(
  raw_list,
  plot_type    = "density",
  view_mode    = "sample",
  metadata     = data_obj$metadata,
  group_column = "Sample Group"
)
```

***

## Step 4 - Normalization

### What methods are available?

```{r norm-methods}
get_normalization_methods()
```

### Compare two pipelines side-by-side

It is good practice to compare normalization strategies visually before
committing to one. Below we compare TIC+Log2 (the most common starting point)
against PQN+Log2 (more robust when large fold-changes are expected).

```{r norm-compare, fig.cap="Side-by-side comparison of two normalization pipelines applied to the same raw data.", fig.height=9}
pipeline1 <- apply_normalizations(data_obj$numeric_data, c("TIC", "Log2"))
pipeline2 <- apply_normalizations(data_obj$numeric_data, c("PQN", "Log2"))

p1 <- create_pipeline_plot(pipeline1,
                            title     = "Pipeline 1: TIC + Log2",
                            metadata  = data_obj$metadata,
                            plot_type = "boxplot")

p2 <- create_pipeline_plot(pipeline2,
                            title     = "Pipeline 2: PQN + Log2",
                            metadata  = data_obj$metadata,
                            plot_type = "boxplot")

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

### Apply the chosen pipeline

```{r normalize}
norm_data <- apply_normalizations(
  data_obj$numeric_data,
  methods = c("TIC", "Log2")
)

dim(norm_data)  # samples x lipids - same dimensions as raw data
```

```{r norm-boxplot, fig.cap="Per-sample distributions after TIC + Log2 normalization. Medians are now aligned across groups."}
norm_list <- list(numeric_data = norm_data, metadata = data_obj$metadata)

visualize_raw_data_improved(
  norm_list,
  plot_type    = "boxplot",
  view_mode    = "sample",
  metadata     = data_obj$metadata,
  group_column = "Sample Group"
)
```

***

## Step 5 - Quality Control with PCA

PCA on normalized data should show samples clustering by biological group
rather than by technical artefact.

```{r pca, fig.cap="PCA of TIC + Log2 normalized data. Ellipses are 95% confidence regions."}
pca_res <- perform_pca(norm_data, data_obj$metadata, "Sample Group")

create_pca_plot_with_ellipses(
  pca_data           = pca_res$pca_data,
  variance_explained = pca_res$variance_explained,
  ellipse_type       = "confidence",
  show_sample_labels = TRUE,
  title              = "PCA - TIC + Log2 Normalized Data"
)
```

***

## Step 6 - Differential Analysis

### Automatic pairwise contrasts

`create_default_contrasts()` generates all pairwise comparisons from the
groups present in your data. You can also specify contrasts manually.

```{r contrasts}
groups    <- unique(data_obj$metadata$`Sample Group`)
contrasts <- create_default_contrasts(groups)

cat("All pairwise contrasts (", length(contrasts), "total ):\n")
print(contrasts)
```

### Run limma

```{r diff-analysis}
diff_res <- perform_differential_analysis(
  data_matrix    = norm_data,
  metadata       = data_obj$metadata,
  group_column   = "Sample Group",
  contrasts_list = contrasts,
  method         = "limma"
)
```

```{r diff-summary}
cat(sprintf("%-28s  %5s  %5s  %4s  %5s\n",
            "Contrast", "Total", "Sig", "Up", "Down"))
cat(strrep("-", 55), "\n")

for (nm in names(diff_res$results)) {
  res    <- diff_res$results[[nm]]
  n_sig  <- sum(res$adj.P.Val < 0.05, na.rm = TRUE)
  n_up   <- sum(res$adj.P.Val < 0.05 & res$logFC > 0, na.rm = TRUE)
  n_down <- sum(res$adj.P.Val < 0.05 & res$logFC < 0, na.rm = TRUE)
  cat(sprintf("%-28s  %5d  %5d  %4d  %5d\n", nm, nrow(res), n_sig, n_up, n_down))
}
```

### Top results from the first contrast

```{r top-results}
first_contrast <- names(diff_res$results)[1]
res_df <- diff_res$results[[first_contrast]]
res_df <- res_df[order(res_df$adj.P.Val), ]

top10 <- utils::head(res_df[, c("logFC", "AveExpr", "adj.P.Val")], 10)
top10$logFC     <- round(top10$logFC, 3)
top10$AveExpr   <- round(top10$AveExpr, 2)
top10$adj.P.Val <- signif(top10$adj.P.Val, 3)

cat("Top 10 features in:", first_contrast, "\n\n")
print(top10)
```

***

## Step 7 - Results Visualization

### Volcano plot

Significant lipids (|logFC| > 1 and adj.P.Val < 0.05) are coloured by lipid
class. The top 10 most significant are labelled.

```{r volcano, fig.cap="Volcano plot coloured by lipid class. Vertical dashed lines mark logFC = +/-1; horizontal dashed line marks adj.P.Val = 0.05."}
create_volcano_plot_labeled(
  results             = diff_res$results[[first_contrast]],
  title               = paste("Volcano:", first_contrast),
  logfc_threshold     = 1,
  pval_threshold      = 0.05,
  top_labels          = 10,
  classification_data = classification,
  color_by            = "LipidGroup"
)
```

### Heatmap of significant features

```{r heatmap, fig.cap="Heatmap of up to 30 significant lipids. Rows (lipids) and columns (samples) are hierarchically clustered. Colour scale is z-scored per lipid.", fig.height=7}
sig_lipids <- rownames(res_df)[
  !is.na(res_df$adj.P.Val) &
  res_df$adj.P.Val < 0.05 &
  abs(res_df$logFC) > 1
]

if (length(sig_lipids) >= 5) {
  top_n  <- min(30, length(sig_lipids))
  avail  <- intersect(sig_lipids[seq_len(top_n)], colnames(norm_data))
  hm_mat <- t(norm_data[, avail, drop = FALSE])

  create_heatmap_robust(
    data_matrix  = hm_mat,
    metadata     = data_obj$metadata,
    group_column = "Sample Group",
    top_n        = top_n,
    title        = paste("Significant lipids:", first_contrast)
  )
} else {
  message("Fewer than 5 significant lipids at these thresholds.")
}
```

***

## Step 8 - Lipid Expression

`create_lipid_expression_barplot()` shows per-sample abundance of individual
lipids, sorted by group. It returns a single ggplot object when one lipid is
selected, or a named list of ggplot objects when multiple lipids are selected.

```{r lipid-expression, fig.cap="Per-sample abundance of the 3 most significant lipids. Samples are automatically sorted by group."}
# Take the 3 most significantly changed lipids
top_lipids <- utils::head(
  rownames(res_df)[!is.na(res_df$adj.P.Val)], 3
)

cat("Selected lipids:\n")
print(top_lipids)

expr_result <- create_lipid_expression_barplot(
  data_matrix     = norm_data,
  metadata        = data_obj$metadata,
  selected_lipids = top_lipids,
  group_column    = "Sample Group",
  data_type       = "normalized"
)

# The function returns a single ggplot (1 lipid) or a named list (multiple)
if (inherits(expr_result, "ggplot")) {
  print(expr_result)
} else {
  for (lipid_plot in expr_result) print(lipid_plot)
}
```

***

## Step 9 - Enrichment Analysis

Enrichment analysis tests whether lipids in a particular class are collectively
more increased or decreased than expected by chance, using the fgsea algorithm.

```{r enrichment}
enrich_res <- perform_enrichment_analysis(
  results_list        = diff_res$results,
  classification_data = classification,
  min_set_size        = 3,
  max_set_size        = 500
)

cat("Contrasts with enrichment results:", length(enrich_res), "\n")
cat("Categories per contrast          :", paste(names(enrich_res[[1]]),
                                                collapse = ", "), "\n")
```

```{r enrichment-table}
# Show top enrichment results for the first contrast, by LipidGroup
grp_enrich <- enrich_res[[first_contrast]][["LipidGroup"]]

if (!is.null(grp_enrich) && nrow(grp_enrich) > 0) {
  grp_enrich <- grp_enrich[order(grp_enrich$pval), ]

  sig <- grp_enrich[grp_enrich$padj < 0.25, ]

  if (nrow(sig) > 0) {
    cat("Significantly enriched lipid groups (padj < 0.25) in:", first_contrast, "\n\n")
    out        <- sig[, c("pathway", "NES", "pval", "padj", "size")]
    out$NES    <- round(out$NES, 3)
    out$pval   <- signif(out$pval, 3)
    out$padj   <- signif(out$padj, 3)
    print(out, row.names = FALSE)
  } else {
    cat("No lipid groups reach padj < 0.25 in this contrast.\n")
    cat("Top 5 by p-value (", first_contrast, "):\n\n")
    out        <- utils::head(grp_enrich[, c("pathway", "NES", "pval", "padj", "size")], 5)
    out$NES    <- round(out$NES, 3)
    out$pval   <- signif(out$pval, 3)
    out$padj   <- signif(out$padj, 3)
    print(out, row.names = FALSE)
  }
} else {
  cat("No enrichment results available.\n")
}
```

```{r enrichment-plot, fig.cap="Enrichment dot plot for the first contrast. Dot size reflects set size; colour reflects statistical significance."}
if (!is.null(grp_enrich) && nrow(grp_enrich) > 0) {
  create_enrichment_dotplot(
    enrichment_data = grp_enrich,
    title           = paste("Lipid Group Enrichment:", first_contrast),
    max_pathways    = 10
  )
}
```

***

## Session Information

```{r session-info}
sessionInfo()
```
